Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-29 analysis/paper/paper.Rmd
working on the discussion section

The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them (Table \@ref(tab:FA-composition-table)). The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds in most potsherds (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of C16:0 and C18:0 fatty acids in our samples suggest that the original organic residues were highly degraded during post-depositional processes. In addition, some potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. A larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005]. Although oleic acid (C18:1) is the most abundant compound for most of our samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish  [@Shoda2018; @Kwak2020]. In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food categories indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by terrestrial animals, particularly ruminant fats or mixtures of terrestrial animals and plants. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might be C3 plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields for all potsherds support that those ceramics were used as cooking or storage pots. One potshard dating to around 4500 years ago demonstrates a rich composition of long-chain fatty acids, suggesting a better condition for preserving lipid molecules. We also recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes from both Guijiabao and Gaoshan. However, the general pattern of the isotopic values of lipid molecules could provide informative clues for a better understanding of the local subsistence of Guijiabao.

Ruminant fats in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, with stone tools associated with hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. However, whether horses were part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value located in ranges of deer or sheep fats based on a study from England [@Evershed2002]. Although the referenced isotopic values were from modern animals and should be ideally adjusted in terms of local environments, they could still provide useful information and have been used as references in many studies []. It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were parts of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress that will allow evaluating our findings.

It is interesting to find a non-ruminant origin that could be C3 plants.

The results show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source. Our striking finding might suggest that consumption of rice could be part of the daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously. Our next steps is to increase sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.

net sinkers
1000 BC horse
Gaushan-90 pig, plant-most are rice: 95% and a few millet  
ecological zones

Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded over the past 4500 years. A large-scale analysis of ceramic across the region may offer insight into regional variations in food choices and the spread of agriculture.

# Conclusion
Our investigation on the ceramics excavated from Southwest China suggests that a mixture of C3 plants and ruminant animals could be the food sources in this region during the Neolithic transition. This case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region. Our case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.

>>>>

The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them (Table \@ref(tab:FA-composition-table)). The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds in most potsherds (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, some potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. A larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005]. Although oleic acid (C18:1) is the most abundant compound for most of our samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.

@@

The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998; @Suryanarayan2021]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020]. In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.

@@

# Discussion
<!--A brief summary of interesting results-->
Our results show that food categories indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by terrestrial animals, particularly ruminant fats or mixtures of terrestrial animals and plants. Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from terrestrial plants. Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots. The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020]. But we also found a base sample dating to ca. 4500 BP. contains a rich composition of long-chain fatty acids. We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes. However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence of Guijiabao.

<!--discussion about possible plant origins-->
One aim of our study is to explore whether new crops, such as C3 plants (eg. rice), C4 plants (eg. millet), or new species (eg. pigs/horses) became an important part of local diet in Yanyuan during the Neolithic transition. We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies [@Shoda2018; @Kwak2020]. Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories. Also, recent macro-botanical and phytolith studies at Guijiabao suggest the presence of a mixing farming of rice and millet, indicating that the origin of C3 plants from our sample could come from rice [@Hao2022 @Huan2022]. It is also possible other grasses classified as C3 plants, such as Pennisetum, were part of the local diet [@Guijiabao2016]. We should note that freshwater fish could also be the origin of non-ruminant fats since C3 plants and freshwater fish have similar isotopic values [@Lucquin2016]. Fish bones and net sinkers were also found at Guijiabao which suggests the possibility of fishing. There are no C4 plants, such as millets, directly detected in our sample, which contradicts the predominance of millet grains found at Guijiabao. But it is hard to make an inference about the role of rice/millet because the results might be caused by small sample effects or sampling strategy.

<!--discussion about possible ruminant fat origin-->
Ruminant fats in the Neolithic potsherds around ca. 4500 BP. from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, with stone tools associated with hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a]. But whether horses were part of local food choices in the Neolithic period is questionable. First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006]. Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004]. The evidence from Guijuabao also confirms a later presence of horse remains. In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value located in ranges of deer or sheep fats based on a study from England [@Evershed2002].  It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016]. A detailed analysis of faunal remains from Guijiabao is still in progress, allowing further evaluation of our findings.

In general, our results suggest that Guijiabao residents relied mostly on ruminants fats with terrestrial plants as supplements. The overall depleted δ^13^C value shows that the diet of ruminants was mainly C3 plants that could come from wild weeds [@Guijiabao2016].

--Gaushan-90 pig, plant-most are rice: 95% and a few millet  
--ecological zones

A large-scale analysis of ceramic across the region may offer insight into regional variations in food choices and the spread of agriculture. Our next step is to increase the sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.

# Conclusion

Our investigation of the ceramics excavated from Southwest China suggests that a mixture of C3 plants and ruminant animals could be the food sources in this region during the Neolithic transition. This case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region. Our case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.

The results show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant sources or marine sources. Our striking finding might suggest that the consumption of rice could be part of daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously.
